---
title: "Tidycensus Walkthrough"
author: "Jackson M Luckey"
date: "3/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadPackages, echo=FALSE}
library(tidyverse)
library(knitr) # for kable
```

# Introduction to Tidycensus

The R package Tidycensus provides a Tidyverse-friendly way of working with census data. Tidycensus was created by Kyle Walker, who has provided some documentation [here](https://walkerke.github.io/tidycensus/articles/basic-usage.html). Tidycensus allows "R users to return Census and ACS data as tidyverse-ready data frames, and optionally returns a list-column with feature geometry for many geographies". The "list-column with feature geometry" allows the R user to easily draw maps using the downloaded census data and ggplot2. Dataframes can be downloaded in either a wide or tidy format, and brief descriptions of each variable are available. The package is an API wrapper for data.census.gov, and the API calls it creates can be accessed and manually ran using packages such as rvest. The package supports both American Community Survey and Decennial data. To pull down data with Tidycensus, you need to know the variable/table name, the year, the geographic level, and the survey.

 To identify the variables and tables available from a survey, use the command `tidycensus::load_variables(year, survey)`, where year is a year in numerical format (e.g. 2018), and survey is a survey in character format (e.g. "acs1"). This returns a dataframe with the columns `name`, `label`, and `concept`. To search for variables within the survey, filter the output of `load_variables()` with `filter(stringr::str_detect(column, pattern))`. If you don't know the name of the table you want to work with, filter on the `concept` column as shown below:
 
```{r searchVariablesByConcept, echo=TRUE}
tidycensus::load_variables(2018, "acs1") %>%
  mutate(concept = stringr::str_to_lower(concept)) %>% # makes searching case insensitive
  filter(stringr::str_detect(concept, "education")) %>%
  slice(1:10) %>%
  kable(caption = "First 10 Variables Related to Education")
```

If you already know the table name you're interested, filter on the `name` column and set pattern equal to the table name followed by an underscore. For example, the American Community Survey one year estimates stores disability data in the table `B18101`. To look at the variables in that table, filter the results of `tidycensus::load_variables() as shown below:

```{r}
tidycensus::load_variables(2018, "acs1") %>%
  filter(stringr::str_detect(name, "B18101_")) %>%
  kable(caption = "Disability Variables in Table B18101")
```

In every case, the `tidycensus::load_variables()` function will return half of the number of variables that the actual census query will return. This is because each variable comes with both an estimate and margin of error.

The geographical level, survey, and year are all easier to determine. Geographical levels are things like state, county, school district, census tract, and congressional district. They are represented as characters. Not all geographical levels are supported by Tidycensus, and not all surveys are available for all geographies. In general, surveys collected less are more likely to include smaller geographical levels. Surveys include the different American Community Surveys, the Decennial census, and supplemental estimates. They can be found on the U.S. Census's website. Finally, year is simply a valid year for that survey provided as a number. Make sure to remember that variables can change between survey years.

To pull down actual data from census.data.gov using Tidycensus, you use one of three functions depending on the survey that you are working with. To work with the decennial census, use `get_decennial()`. To work with American Community Survey data, use `get_acs()`. Finally, use `get_estimates()` to work with the Census Bureau's population estimates API. If you download the data in wide format, Tidycensus returns a table with the columns `GEOID` and `NAME`, which refer to the geographical region, and two columns, `Table_Variable + E` and `Table_Variable + M`, for each variable in the table. For example, if the table was "someTable", and it only included the variables "001" and "002", the returned dataframe would included the columns `someTable_001E, someTable_001M, someTable_002E, someTable_002M`. E refers to estimate, while M refers to margin of error. Finally, if you included the argument `geometry = TRUE` in the call, there will be a column `geometry` that includes the data required to draw a map of the geographical regions included in the call.

In order to better use Tidycensus, I wrote custom functions to accomplish several repeated tasks. One function, `get_census_tables_multiple_years(table, years)`, allows the user to download multiple years of a particular census table at once. The functions accept a table code and year range, and downloads the table for each year, pivot it into a wide format, adds a year column, and binds the rows of the resulting dataframes into one master dataframe, which is returned. Another function, `clean_county_col(df)`, modifies the "county" column, removing the string "County, Ohio" from each entry in order to make it easier to join with other datasets.